### 2008-05-30 trunk/Papers/rmrs-08/entailment.tex
very prelim version

Now that we have the formal tools in place, let's go back to the
problem of checking compatibility of semantic outputs.  We can phrase
this as RMRS entailment.  This is progress already.

Then show the representation theorem of entailment as solved form
comparison:

\begin{thm}
  Let $\varphi, \varphi'$ be two RMRSs.  Then $\varphi \models
  \varphi'$ iff for every solved form $S$ of $\varphi$, there is a
  solved form $S'$ of $\varphi'$ such that $S$ is an extension of
  $S'$.
\end{thm}

(Of course, this doesn't give us an efficient algorithm for entailment
checking, because the number of solved forms may be exponential on
each side.  And we still don't know the complexity of deciding
entailment (even for normal dominance constraints).  Nevertheless,
this is nice, because it illustrates that our model theory is both
pretty and useful for proving stuff.)

>>>>

As discussed in Sections~\ref{sec:intro} and~\ref{sec:motivation}, 
{\sc rmrs}'s design aims to make it possible to compare
the semantic output of several language processors of varying depth,
and even combine them when appropriate.  The following theorem 
demonstrates that {\sc rmrs}-entailment---defined in terms of
models and or solved forms---provides the logical 
basis for such parser integration:

% NOTE: I'm fudging that this applies only to normal dominance
% constraints that are overlap-free.  We should put this back if there
% is room!
%
\begin{thm}\label{thm:big-one}
  Let $\varphi, \varphi'$ be two {\sc rmrs}s.  Then $\varphi \models
  \varphi'$ iff for every solved form $S$ of $\varphi$, there is a
  solved form $S'$ of $\varphi'$ such that $S$ is an extension of
  $S'$.
\end{thm}

This theorem doesn't provide 
efficient
algorithms for entailment checking, and of course this would be
required to make parser comparison and integration with {\sc rmrs}
practical.  Computing the solved forms of $\varphi$ and $\varphi'$ may
be exponential, and the complexity of the {\sc rmrs} validity problem
is still unknown.  However, the above theorem illustrates that our
model theory for {\sc rmrs} functions in the way it should for parser
comparison and integration, and it can thus be used to prove important
logical properties of shallow parsing systems that adopt {\sc rmrs}.

The outline proof of Theorem~\ref{thm:big-one} is as follows:
\begin{proof}
  ``$\Leftarrow$'': Let $M,\alpha$ an arbitrary model and variable
  assignment that satisfy $\varphi$.  There is a solved form $s$ of
  $\varphi$ such that $M,\alpha \models s$.  By assumption, this means
  that there is a solved form $s'$ of $\varphi'$ such that $s$ is an
  extension of $s'$.  Therefore, $M,\alpha \models s'$ and hence
  $M,\alpha \models \varphi'$.

  ``$\Rightarrow$'': Assume that $\varphi \models \varphi'$, and let
  $s$ be an arbitrary solved form of $\varphi$.  We need to show that
  we can remove atoms from $s$ such that $s$ remains an extension of
  the resulting $s'$ and $s'$ is a solved form of $\varphi'$.

  We construct $s'$ as the conjunction of all atoms in $s$ that only
  contain variables occurring in $\varphi'$ and all atom $X \dom Y$
  such that $X$ and $Y$ occur in $\varphi'$ and $(X,Y) \in D(s)$.

  It is obvious that $s$ is an extension of $s'$.  Furthermore, $s'$
  is in solved form, because $s$ was a tree already and we didn't
  introduce new edges.  The tricky part is to show that $s'$ is a
  solved form of $\varphi'$.  $s'$ contains all labeling atoms of
  $\varphi'$ according to
  Lemma~\ref{lem:entailment-preserves-labeling-atoms}; a similar
  argument holds for inequality atoms.  Finally, $D(\varphi')
  \subseteq D(s')$ because $M,\alpha \models s'$ and $s'$ is
  tree-shaped and normal and therefore contains all statements about
  dominance that are true in $M$.
\end{proof}

\begin{lemma} \label{lem:entailment-preserves-labeling-atoms}
  Under the assumptions of
  Theorem~\ref{thm:big-one}, if $\varphi \models
  \varphi'$, then $\varphi'$ only contains labeling atoms that also
  occur in $\varphi$.  
\end{lemma}
\begin{proof}
  Let $\varphi \models \varphi'$, but let $\varphi'$ contain a
  labeling atom $X \Neq f(X_1,\ldots,X_n)$ that doesn't occur in
  $\varphi$.  Call this atom $A$.

  Furthermore, let $M,\alpha \models \varphi$ arbitrary.  Because
  $M,\alpha \models \varphi'$, we must have $M(\alpha(X)) = f$.

  Now replace the label of $\alpha(X)$ in $M$ by some other symbol $g$
  of arity $n$.  Call the resulting model $M'$.  We still have
  $M',\alpha \models \varphi$ because $\varphi$ doesn't contain a
  labeling atom for $X$.  But now $M',\alpha \not\models \varphi'$.
  This contradicts the entailment assumption.
\end{proof}

It is really necessary to state
Theorem~\ref{thm:big-one} in terms of extensions of
solved forms.  It is \emph{not} true that $\varphi \models \varphi'$
implies that $\varphi$ is an extension of $\varphi'$.  A
counterexample is shown in Fig.~\ref{fig:counterexample}, where the
right-hand constraint entails the left-hand constraint (they are
equivalent), but is not an extension of it.

\begin{figure}
  \centering
  \includegraphics[scale=0.6]{pic-extension-counterexample}
  \caption{A counterexample to a simpler version of Theorem~\ref{thm:big-one}.}
  \label{fig:counterexample}
\end{figure}

### 2008-05-30 trunk/Papers/rmrs-08/introduction.tex
very prelim version

\section{Introduction} \label{sec:introduction}

What we'll do here.  Problem: integrate semantic representations
computed by deep and shallow parsing.  We represent both in RMRS,
which we define cleanly here for the first time.  We also introduce a
model theory for RMRS.  This allows us to define ``compatibility'' of
deep and shallow semantic representations as entailment of RMRS
representations; we define entailment and characterize it
syntactically as extension of solved forms.

just so we have a citation here: \cite{Copestake_etal:MRS}

>>>>

\section{Introduction} \label{sec:intro}

% What we'll do here.  Problem: RMRS is designed to support building
% semantic components to shallow parsers, and is intended to offer a way
% of integrating semantic representations
% computed by deep and shallow parsing.  We demonstrate the feasibility
% of this for the first time, by introducing a model theory for RMRS,
% defining entailment among RMRS representations and characterising this
% entailment relation 
% syntactically as extension of solved forms (ref to solved forms).  
% We show that the ``compatability'' of
% deep and shallow semantic representations is definable in terms of the
% entailment 
% relation between the RMRS
% representations.

% just so we have a citation here: \cite{Copestake_etal:MRS}

Representing semantics as a logical form that supports automated
inference and model construction is vital for deeper language
engineering tasks, such as dialogue systems.
These need access to the vast array of
semantic information that is captured in conventional logical forms,
including (but not limited to) information about semantic scope and
predicate argument structure.  Hand-crafted grammars yield detailed
logical forms (e.g., \cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing tends to lack
robustness: hand-crafted grammars fail to cover all words and
linguistic constructions; and unedited text and speech contains
ill-formed phrases, which by design deep grammars do not handle.  
\hidden{
Deep
parsers on their own also lack ambiguity resolution techniques,
producing all possible analyses.  Integration of statistical parse
selection with deep parsing is promising (e.g.,
\cite{toutanova:etal:2004}), but is not a solved problem.
}

Robost language processors that produce a single conventional logical
form for a given natural language string are beginning to emerge
(e.g.,
\cite{bos:etal:2004,rupp:etal:2000,wong:mooney:2006,zettlemoyer:collins:2005}).
But the output of these systems don't relate to any gold standard deep
parse as produced by expert grammar developers (for instance, while
the training corpus used in \cite{zettlemoyer:collins:2005} features
control phenomena in the language strings, their logical forms don't
represent it).  This makes it hard to judge the logical forms that the
models derive from a linguistic perspective; nor can one integrate
their output with that of a hand-crafted grammar when desired.

This paper focusses on a particular approach to producing semantic
information from robust parsers, exemplified in
\cite{copestake:2003,frank:2004}, among others.  Their strategy is to
utilise semantic underspecification to semi-automatically build
semantic components to shallow parsers, so that the output neither
over-determines nor under-determines the semantic information that is
revealed by the (shallow) syntactic analysis.  The semantic formalism
used to express this is Robust Minimal Recursion Semantics ({\sc
  rmrs}, \cite{copestake:2003}); this is an extension of {\sc mrs}
\cite{copestake:etal:2005} that is designed to be maximally flexible
in the type of semantic information that can be left underspecified:
it can express partial information about semantic scope, the values of
arguments to predicate symbols and/or their argument position, the
arity of the predicate symbols and the sorts of arguments they take.
We show in Section~\ref{sec:motivation} that all these features are
needed when information about lexical subcategorisation or
syntactic dependencies is missing---a characteristic feature of
shallow parsers.  Several researchers have demonstrated that {\sc
  rmrs} is a suitable framework on which to semi-automatically
construct semantic components to shallow parsers, ranging in depth
from {\sc pos} taggers (refs) to chunk parsers (frank ref) and
intermediate parsers ({\sc rasp} ref).

A major motivation for adopting {\sc rmrs} over other techniques for
robustly deriving logical forms is the promise that it can form the
basis for integrating the output of several parsers, and be compared
in particular with the output of a hand-crafted grammar.  This paper
demonstrates the feasibility of this integration for the first time,
by introducing a model theory for {\sc rmrs}, that in turn defines
entailment among {\sc rmrs} representations.  This entailment relation
is also characterised syntactically as an extension of solved forms
(ref to solved forms).  We show that the proof theory and model theory
of {\sc rmrs} that results provides a formal basis for integrating the
semantic output of several shallow parsers, for checking the
satisfiability of a shallow parse, and for testing its compatibility
with a deep parse.

### 2008-05-30 trunk/Papers/rmrs-08/motivation.tex
very prelim version

Motivate what we're doing and provide background.  Example outputs of
deep and shallow parsers.  Illustrate and state the problem.

>>>>

% Motivate what we're doing and provide background.  Example outputs of
% deep and shallow parsers.  Illustrate and state the problem.
%
% Try to find an eg where the comparison isn't easy to see (from Anna
% Ritchie's deliverable?). 
% NOTE: on Ritichie: Perhaps "Three dogs bark on Tuesday in
% June" from page 11?  It's hard to follow what's going on, from the
% sheer size of the RMRSs and the need to allow for variation on the
% variable names.  But I don't know how to extract this figure
% techinically; nor whether we have room.

We motivate and describe the main features of {\sc rmrs} with a (toy)
example.  Sentence 
(\ref{eg:cat}a) features a semantic scope ambiguity that, according to
most deep grammars, syntax
doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
reflects this by underspecifying socpe: (i) each predication is
labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); and (iii)
constraints on scope are expressed by the qeq conditions (roughly, $h
=_q l$ means that only zero or more quantifiers lie between the scopal
argument $h$ and the predication labelled by $l$ in any fully-resolved
logical form); and (iv) these scope
constraints allow for more than one way of `plugging' the holes with
labels.\footnote{The event variable in the predicate symbol
  corresponding to the adjective is motivated by copulas---{\em The
    cat is fat} arguably has the logical form 
  $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
\begin{examples}
\item   \label{eg:cat}
\begin{subexamples}
\item
Every fat cat chased some dog.
% \item
% $\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x) \wedge
% \mbox{\_cat\_n\_1}(x),$\\
% \hspace*{0.5in}$\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),
% \mbox{\_chase\_v\_1}(e_{spast},x) )$\\
% $\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),$\\
% \hspace*{0.5in}$\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x)
% \wedge 
% \mbox{\_cat\_n\_1}(x), 
% \mbox{\_chase\_v\_1}(e_{spast},x)))$
\item   $l_1\handel\mbox{\_every\_q\_1}(x,h_2,h_3)$\\
$l_4\handel\mbox{\_fat\_j\_1}(e',x)$\\
$l_4\handel\mbox{\_cat\_n\_1}(x)$\\
$l_5\handel\mbox{\_some\_n\_1}(y,h_6,h_7)$\\
$l_8\handel\mbox{\_dog\_n\_1}(y)$\\
$l_9\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
$h_0 =_q l_9, h_2 =_q l_4, h_6 =_q l_9$
\end{subexamples}
\end{examples}
(\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
conjunction through label sharing (see $l_4$ above), and the {\sc
  erg}'s naming convention for predicate symbols: each predicate is
constructed from the word lemma, its syntactic category and a sense
number, and the leading underscore identifies lexical predicates.

Conceptually, this {\sc mrs} is a partial description of all its
possible (fully specific) interpretations in context.  
These are expressed in a so-called {\em base language}, where each
predicate symbol in the {\sc mrs} is a proxy for one or more
base-language constructors of the same arity
Assuming a very close
correspondence between the {\sc mrs} predicate symbols and the
base-language constructors they 
stand for, the {\sc mrs}
(\ref{eg:cat}b) (partially) describes both the base-language formulae in
Figure~\ref{fig:1}, shown as trees to emphasise that an {\sc mrs} is a
partial description of the formulae's {\em form}; not its
(base-language) entailments. 
In essence, the labels in the {\sc mrs} reify the
nodes in the tree, and each of these trees encodes one way of equating
holes with labels that respects qeq constraints and binding conditions
on the variables.

\begin{figure}[h]

Put trees in here

\caption{The Form of alternative base logical forms of
  (\ref{eg:cat}a).}
\label{fig:1}
\end{figure}

The {\sc mrs} (\ref{eg:cat}c) is a notational variant of the {\sc
  rmrs} (\ref{eg:cat}d). {\sc rmrs}s offer a more factorised
representation: its predicate symbols are all unary and the other
arguments to the (base-language) constructor that it corresponds to
are represented by separate binary relations on the unique {\em
  anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
variable and label equalities (e.g., $x=x_1$, $l_{41}=l_{42}$).
\begin{examples}
\item   [\ref{eg:cat}]
\begin{subexamples}
\item   [c]
$l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
\hspace*{0.1in}$\mbox{RSTR}(a_1,h_2),
\mbox{BODY}(a_1,h_3)$\\ 
$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \mbox{ARG1}(a_{41},x_2)$\\
$l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
$l_5:a_5\handel\mbox{\_some\_q\_1}(x_4)$,\\
\hspace*{0.1in}$\mbox{RSTR}(a_5,h_6),
\mbox{BODY}(a_5,h_7)$\\ 
$l_8:a_8\handel\mbox{\_dog\_n\_1}(x_5)$\\
$l_9:a_9\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
\hspace*{0.1in}$\mbox{ARG1}(a_9,x_6),
\mbox{ARG2}(a_9,x_7)$\\ 
$h_2=_q l_{42}, l_{41}=l_{42}, h_6 =_q l_8, h_0=_q l_9$\\
$x_1=x_2,x_2=x_3, x_4=x_5,$\\
$x_3=x_6, x_5=x_7$
\end{subexamples}
\end{examples}

This factored representation allows one to build semantic components
to shallow parsers where lexical or syntactic information is absent.
An extreme example would be a {\sc 
  pos} tagger: one can build its semantic component simply by deriving
lexical predicate symbols from the word lemmas and their tags and ensuring
that the predication for each word receives a unique label, anchor and
argument, as given in (\ref{eg:cat-pos}):
\begin{examples}
\item   \label{eg:cat-pos}
\begin{subexamples}
\item   Every\_{\sc at1} fat\_{\sc jj} cat\_{\sc nn1} chased\_{\sc
    vvd} some\_{\sc dd} dog\_{\sc nn1}
\item   $l_1:a_1:\mbox{\_every\_q}(x)$, \\
$l_{41}:a_{41}:\mbox{\_fat\_a}(e')$,\\
$l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
$l_9:a_9:\mbox{\_chased\_v}(e_{past})$, \\
$l_5:a_5:\mbox{\_some\_q}(x_4)$, \\
$l_8:a_8:\mbox{\_dog\_n}(x_5)$
\end{subexamples}
\end{examples}
Semantic relations, sense tags and the arity of the base-language
constructors are
missing from (\ref{eg:cat-pos}b) because the {\sc pos} tagger doesn't
reveal information about syntactic constituency, word sense or lexical
subcategorisation respectively.  
But the {\sc rmrs}s (\ref{eg:cat}c) and
(\ref{eg:cat-pos}b) are entirely compatible, the former being more
specific than the latter.  At least, this holds so long as the
base-language constructors that the {\sc rmrs} predicate symbol
{\_name\_t\_s} corresponds to is a 
subset of those corresponding to the predicate symbol {\em \_name\_t},
for any word lemma {\em name} and {\sc pos} tag $t$ (written 
$\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).

The values of non-scopal arguments are underspecified through the
absence of ARG relations and/or through the absence of
equalities and inequalities among the variables $x_1, x_2,\ldots$.
The need to
underspecify such semantic dependencies is required when
the shallow parser doesn't record syntactic dependencies.
Indeed, there is no single {\sc mrs} which is satisfied by both the
trees in Figure~\ref{fig:2}, assuming of course that disjunction is not part of
the {\sc mrs} language.  In words, these trees collectively specify
the (partial) semantic information that $x$ is either the second
argument (i.e., the direct object) or the third argument (i.e., the
indirect object) to {\em give}, but we don't know which.  
A shallow
processor that misses long distance dependencies in
(\ref{eg:which-dog}) cannot
discern that the bone is the second argument to {\em
  give} in (\ref{eg:which-dog}a), while the dog is the third argument
in (\ref{eg:which-dog}b).
\begin{examples}
\item   \label{eg:which-dog}
\begin{subexamples}
\item   To which dog did Kim give the bone?
\item   Which bone did Kim give the dog?
\end{subexamples}
\end{examples}
Thus to maximise the semantic contribution
of the parser while not overdetermining it, one should express that
the argument position of the noun phrases in {\em give the bone} vs.\
{\em give the dog} are underspecified between the second and third
position, as shown in Figure~\ref{fig:2}.   This will be expressed in
{\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$
between the anchor $a$ of the predicate symbol for {\em give} and the
individual $x$.   More generally, $\mbox{{\em ARG}}_n(a,x)$ means that
$x$ is an argument to the predicate anchored at $a$, but its argument
position is unspecified.

\begin{figure}

\leaf{$e$}
\leaf{$\ldots$}
\leaf{$x$}
\leaf{$\ldots$}
\branch{4}{\_give\_v\_b1}
\tree
\hfill
\leaf{$e$}
\leaf{$\ldots$}
\leaf{$\ldots$}
\leaf{$x$}
\branch{4}{\_give\_v\_b1}
\tree

\caption{$x$ is a participant in a
  giving event, but $x$'s role is unknown.}
\label{fig:2}
\end{figure}

The binary ARG relations also supports underspecifying the {\em arity}
of the base-language constructors, as is required when lexical
subcategorisation information is missing from the shallow parser.
While a deep parser assumes complete information about lexical
subcategorisation in any of its derivations, thereby fixing the arity
of the base-language constructors in the fully-specific logical form
to be the same as the {\sc mrs} predicate symbol $P$ that's introduced by
the lexeme, the predicate symbols in an {\sc rmrs} yielded from a
shallow parse should, in principle at least,
correspond to base language constructors of mixed arity.  For
instance, the English verb {\em bake} can be used causatively ({\em
  Kim baked the potatoes}), in which case semantically it corresponds
to a 3-place predicate symbol {\em \_bake\_v\_b1} in the base
language, or it may be used 
inchoatively ({\em the potatoes are baking}), that semantically
corresponds to a 2-place predicate {\em \_bake\_v\_b2}.  So the 
predicate symbol {\em 
  \_bake\_v} in the {\sc rmrs} that's introduced by a {\sc pos} tagger
must be able to denote either of these
base-language predicates in the {\sc rmrs}'s interpretation.

\hidden{
Finally, {\sc rmrs} allows one to underspecify the {\em sort} of an
argument by including in its language meta-variables $u_1,u_2,\ldots$
over individual, event, and handle variables.  For instance, the predicate
symbol {\em \_be\_v} produced by a {\sc pos} tagger is resolvable to several
3-place base-language constructors whose argument-types 
vary: the base-language constructor corresponding
to its use in {\em John is a man} takes an event and two individuals
as arguments; but in {\em It is John who talked} the constructor takes
an event, an individual and a proposition as arguments.
}
